### 2024-03-10 thesis/abstract.tex
updated thesis project
6 march
March 10th 2024

\begin{abstract}
    The theory of kernel methods will be applied to the problem of probabilistic forecasting taking as input some data.
    Conceptually, the proposed analysis could be applied to any kind of data. In this study we considered the electricity market, because of the interesting implications  on risk management tasks.

>>>>

% PRELIMINARY ABSTRACT, ONCE I AM DONE WRITE THE FINAL ABSTRACT
\begin{abstract}
    The theory of kernel methods will be applied to the problem of point and probabilistic forecasting for the energy sector.
    Such choice is motivated by its interesting implications.

    % on risk management tasks.

    % - Contribution: Kernel herding applied to the
    % field of probabilistic forecasting

### 2024-03-10 thesis/input_files/literaturereview.tex
March 10th 2024

Electricity Forecasting 
%intro
During the past 25 years %because paper of 2014 says 15, so now it is 25
a wide range of new ideas have been proposed for point forecasting and for probabilistic forecasting.
\\
The field benefitted greatly from the increse of computing power, the greater availability of dataset and the interest in data science.
As a consequence, the forecaster's toolbox has grown in size and complexity.
\\
As already pointed out, such variety of methods is characterized by the heterogeneity in the fields from which they come from; methods come from statisitics, mathematics, econometrics, electrical engineering and the artificial intelligence communities.
\\
Before delving into the literature review, it is important to make clear that at this point in time there is no superior method. Different solutions may outperform or underperform compared to other techniques depending on the problem settings. Thus, understanding the complexity, strenghts and weaknesses of each method is crucial for fitting the right model to the right setting.
\\
Finally, within this research community, it emerged the need of more homogeneity in the choice of the error valuation metrics (section \ref{metrics}), data quality and in the way of comparing model performances \cite{EPF_review}; as a solution, \cite{review_2021} proposes a checklist to aid evaluating the meaningfulness of new research. Hence, throughout this thesis work, we will stick to the proposed principles and best practices peculiar of the EF field.
%literature review

@@

\subsection{Point vs Probabilistic}\label{cls_pp}
Finally, EF literature distinguishes between point and probabilistic forecasts. Each of the two has its advantages and disadvantages; point forecasts are easier to generate and less computationally intensive while probabilistic forecast are more informative. Industry and research efforts have focused primarily on point forcasting. Nevertheless, interest in probabilistic forecasting has risen considerably over the last years due to renewable integration requirements, introduction of smart grids and increased market competitivness.
\\
\\
To conclude this section
Ziel et Steinert carefully tables, are reported in the appendix for the interested reader.

@@

This section presents the results of the bibliometric analysis we performed on 6th March 2024. This survey has been carried out by using the Scopus citation database. For details on the specific queries entered in Scopus, refer to \ref{src_code}.

@@

Finally, in order to refer to the most relevant source in the field, EF outlets have been ranked by popularity and plotted in \ref{fig:src_title}.

@@

He et al. \cite{he2022cooperative} models  multistep wind speed probabilistic forecasting by mixing CEEDMAN, LASSO and QRNN. In this work, CEEMDAN is used to decompose the wind speed time series; LASSO compresses high dimensional features; QR is used for obtaining quantile forecasts; finally KDE converts quantile forecasts into density estimates.
Wan et al. \cite{wan2016direct} proposes a combination of QR and ELM to generate non parametric probabilistic forecast of wind generation.

@@

Marcjasz et al \cite{probablistic_electricity_forecast2} uses distributional neural networks to crete probabilistic forecasts for the day ahead electricity prices in the german market.

@@

Zhang et al. \cite{zhang2020probability} introduce a framework based on quantile regression and kernel density estimation in the context of short term wind forecasting. The proposed methods behave well compared to a ARMA(1,1).

@@

@@

Huurman et al. \cite{huurman2012power} surveys the predictive power of weather variables for electricity prices in danish market. Their empirical results suggest that weather is central for point forecasting day ahead prices. The opposite conclusion are drawn for density forecasting

@@

Considerations that can be drawn from the above literature review.
Every paper uses different datasets (heterogeneous)
So it is not possible to compare directly results
from one paper to another without implementing the
paper specific algorithms and the applying them to
your dataset.
This is why sections after are destined to analyzing
how these proposed methods so far work and their mathematical
theory details

@@

************************

@@

************************
Lately, the idea of combining forecasts has gained popularity in the forecasting communit \cite{forecasting_big}; in the literature, combined forecasts are called ensemble \cite{gneiting_weather_ensemble}.
Note that the more the errors of the combined models are not correlated the more we can benefit from ensembles.
It is also worth noting that older and simpler methods are still valuable(in combination with other models or on their own); these being less subject to overfitting than complex models.
\section{Kernel methods literature review}
%INTRO KERNEL METHODS E ENERGY PRICE MODELLING
Kernel methods are a class of algorithms for patter analysis.
With kernel methods we are able to apply linear methods with predictors in a high dimensional space, without having to explicitly evaluate the involved dot products of the features.
Throughout this thesis work, we will address the performance of kernel methods in the context of EF.
- Kernel history evolution.
Their name comes from the german word kern, which translates to core in english. Such term was first used by David Hilbert in his paper on integral equations \cite{hilbert} where he defined a definite kernel. Following, Hilbert's and Schmidt's \cite{schmidt} work lead to the introduction of a new space, the Hilbert space.
In 1909, James Mercer improved Hilbert's work by proposing his theorem \cite{mercer}. This theorem underlies the power of kernel methods, that is the kernel trick.
In 1938, Schoenberg \cite{schoenberg} developed the mathematical results that allow us to find the kernel associated to a specific feature space metric.
In 1941, Kolmogorov \cite{kolmogorov} carried out stuides on representing kernel in linear spaces.
In 1950, Aronszajn \cite{aronszajn} published the first work on RKHS; developing the general method for representing kernels in linear spaces.
In 1964, Aizerman \cite{aizerman} further improved the theory of RKHS.
It was in the nineties that theory of kernel methods got popular, particularly in the field of machine learning. Kernels have been used in various different tasks such as SVM \cite{vapnik1} \cite{vapnik2}, Gaussian process classifiers \cite{williams}, spline methods \cite{wahba}, neural networks \cite{poggio} and principal component analysis \cite{pca_scholkopf}.
Nevertheless kernel methods received very little attention in the specific setting of EF literature.
The theory needed for this thesis work is covered in section \ref{kernel_theory}. 
For an introduction to kernel methods, we referred to \cite{learning_with_kernels}.
Kanagawa et Fukumizu introduces to the concept of kernel mean embedding \cite{pmlr}. Muandet et al. \cite{Muandet_2017} surveys established results and new advances in the theory of Hilbert space distribution embeddings. 
It has to be said that computing and storing such embeddings becomes prohibitive for large scale settings. Rudi et al. \cite{2022nystrom} proposes an efficient approximation procedure based on the Nyström method \cite{nystrom}, providing also an upper bound for the approximation error.
Article \cite{supersamples} presents kernel herding; basically, Smola et al. used the kernel trick to extend the herding algorithm to continous spaces. The result is an infinite memory deterministic process that takes in a collection of samples and learns to approximate a pdf.
%QUESTO DA VEDERE
% Particularly extending the idea of \cite{2022nystrom}, where the Nyström approximation is employed in computing the kernel mean embedding, experiments with the Pivoted Cholesky decomposition will be performed.

>>>>

%Electricity Forecasting 
%intro
During the past 25 years %because paper of 2014 says 15, so now it is 25
a wide range of new ideas have been proposed for electricity point forecasting and for probabilistic forecasting.
\\
The field benefitted greatly from the increse of computing power, the greater availability of dataset and the interest in data science.
As a consequence, the forecaster's toolbox has grown in size and complexity.
\\
Before delving into the literature review, it is important to make clear that at this point in time there is no superior method. Different solutions may outperform or underperform compared to other techniques depending on the problem settings. Thus, understanding the complexity, strenghts and weaknesses of each method is crucial for fitting the right model to the right setting.
\\
Finally, within this research community, it emerged the need of more homogeneity in the choice of the error valuation metrics (section \ref{metrics}), data quality and in the way of comparing model performances \cite{EPF_review}; as a solution, \cite{lago} proposes a checklist to aid evaluating the meaningfulness of new research. Hence, throughout this thesis work, we will stick to the proposed principles and best practices peculiar of the EF field.
%literature review

@@

\subsection{Point vs Probabilistic}\label{cls_pp}
Finally, EF literature distinguishes between point and probabilistic forecasts. Each of the two has its advantages and disadvantages; point forecasts are easier to generate and less computationally intensive while probabilistic forecast are more informative. Industry and research efforts have focused primarily on point forcasting. Nevertheless, interest in probabilistic forecasting has risen considerably over the last years due to renewable integration requirements, introduction of smart grids and increased market competitivness.

@@

This section presents the results of the bibliometric analysis we performed on March 6th 2024. This survey has been carried out by using the Scopus citation database. For details on the specific queries entered in Scopus, refer to \ref{src_code}.

@@

Finally, in order to refer to the most relevant source in the field, EF outlets have been ranked by popularity and plotted in figure \ref{fig:src_title}.

@@

He et al. \cite{he2022cooperative} models  multistep wind speed probabilistic forecasting by mixing CEEDMAN, LASSO and QRNN. In this work, CEEMDAN is used to decompose the wind speed time series, LASSO compresses high dimensional features, QR is used for obtaining quantile forecasts, finally KDE converts quantile forecasts into density estimates.
Wan et al. \cite{wan2016direct} proposes a combination of QR and ELM to generate non parametric probabilistic forecasts of wind generation.
Zhang et al. \cite{zhang2019wind} forecasts wind speed by adopting QRMGM, which is a combination of QR with a minimal gated memory network.
Hyndman et al. \cite{hyndman1996estimating} explains kernel estimator of conditional density, analyses its asymptotic behaviour and covers an application for the daily temperatures of Melbourne.
Kaur et al. \cite{kaur2022energy} carries out a comparative study of techniques spanning ideas from statistic and artificial intelligence.

@@

Marcjasz et al. \cite{probablistic_electricity_forecast2} uses distributional neural networks to create probabilistic forecasts for the day ahead electricity prices in the german market.

@@

Zhang et al. \cite{zhang2020probability} introduce a framework based on quantile regression and kernel density estimation in the context of short term wind forecasting. The proposed methods behave well compared to an ARMA(1,1).

@@

%Metrics Start
Koochali et al. \cite{koochali2022random} reviews various existing methods for assessing probabilistic forecast models and discusses their advantages and disadvantages.
Matheson et al. \cite{matheson1976scoring} develops classes of scoring rules for continuous probability distributions.
Gneiting et al. \cite{gneiting2007strictly}
provides a thorough overview of the theory of proper scoring rules for interval and density forecasts.
Gneiting et al. \cite{gneiting2014probabilistic}
covers theory and techniques state of the art in probabilistic forecasting.
%Metrics End

@@

Fatema et al. \cite{fatema2023probabilistic} considers gaussian process regression for point forecasting and prediction intervals prediction. Then, it inputs PI to KDE in order to estimate a probability distribution.
Dudek \cite{dudek2018probabilistic} proposes a probabilistic forecasting model based on the Nadaraya Watson estimator \cite{nadaraya1964estimating} \cite{watson1964smooth}.
Huurman et al. \cite{huurman2012power} surveys the predictive power of weather variables for electricity prices in the danish market. Their empirical results suggest that weather is central for point forecasting day ahead prices. The opposite conclusion are drawn for density forecasting.
Lately, the idea of combining forecasts has gained popularity in the forecasting community \cite{forecasting_big}; in the literature, combined forecasts are called ensemble \cite{gneiting_weather_ensemble}.
Note that, the more the errors of the combined models are not correlated the more we can benefit from ensembles.
It is also worth noting that older and simpler methods are still valuable (in combination with other models or on their own); these being less subject to overfitting than complex models.

@@

A couple of considerations can be drawn from the above literature review.
As already pointed out, the EF field is characterized by heterogeneity in its forecasting techniques; methods come from statisitics, mathematics, econometrics, electrical engineering and the artificial intelligence communities.

@@

Every paper uses different datasets.
Therefore, it is not possible to compare directly results from one paper to another without implementing the paper specific algorithms and then applying them to your dataset. An additional hurdle is that some datasets are not freely accessible.

@@

Thus, a good understanding of state of the art methods in EF is required to carry out a rigorous comparison between methods.
That is why the following chapters are devoted to summarising the mathematical theory underlying such techniques.

\section{Kernel methods literature review}
%INTRO KERNEL METHODS E ENERGY PRICE MODELLING
Kernel methods are a class of algorithms for patter analysis.
With kernel methods we are able to apply linear methods with predictors in a high dimensional space, without having to explicitly evaluate the involved dot products of the features.
Throughout this thesis work, we will address the performance of kernel methods in the context of EF.
\\
%- Kernel history evolution.
Their name comes from the german word kern, which translates to core in english. Such term was first used by David Hilbert in his paper on integral equations \cite{hilbert} where he introduced the term of definite kernels. Following, Hilbert's and Schmidt's \cite{schmidt} work lead to the introduction of a new space, the Hilbert space.
In 1909, James Mercer improved Hilbert's work by proposing his theorem \cite{mercer}. This theorem underlies the power of kernel methods, that is the kernel trick.
In 1938, Schoenberg \cite{schoenberg} developed the mathematical results that allow us to find the kernel associated to a specific feature space metric.
In 1941, Kolmogorov \cite{kolmogorov} carried out stuides on representing kernel in linear spaces.
In 1950, Aronszajn \cite{aronszajn} published the first work on RKHS; developing the general method for representing kernels in linear spaces.
In 1964, Aizerman \cite{aizerman} further improved the theory of RKHS.
It was in the nineties that theory of kernel methods got popular, particularly in the field of machine learning. Kernels have been used in various different tasks such as SVM \cite{vapnik1} \cite{vapnik2}, Gaussian process classifiers \cite{williams}, spline methods \cite{wahba}, neural networks \cite{poggio} and principal component analysis \cite{pca_scholkopf}.
Nevertheless kernel methods received very little attention in the specific setting of EF literature.
\\
The kernel theory needed for this thesis work is covered in section \ref{kernel_theory}. 
For an introduction to kernel methods, we referred to \cite{learning_with_kernels} \cite{hofmann2006review}.
\\
Kanagawa et Fukumizu introduces to the concept of kernel mean embedding \cite{pmlr}. Muandet et al. \cite{Muandet_2017} surveys established results and new advances in the theory of Hilbert space distribution embeddings. 
It has to be said that, computing and storing such embeddings becomes prohibitive for large scale settings. Rudi et al. \cite{2022nystrom} proposes an efficient approximation procedure based on the Nyström method \cite{nystrom}, providing also an upper bound for the approximation error.
Article \cite{supersamples} presents kernel herding; basically, Smola et al. used the kernel trick to extend the herding algorithm to continous spaces. The result is an infinite memory deterministic process that takes in a collection of samples and learns to approximate a pdf.
\\
%QUESTO DA VEDERE
% Particularly extending the idea of \cite{2022nystrom}, where the Nyström approximation is employed in computing the kernel mean embedding, experiments with the Pivoted Cholesky decomposition will be performed.

### 2024-03-10 thesis/input_files/problemdescription.tex
9 march 2024
March 10th 2024

Individuals and organizations constantly face situations of uncertainty. Thus, the need of robust forecasting methods. Such methods are crucial to the process of taking informed decision and to strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guess about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting to the energy sector. 
\\
\\
The reason of our decision to focus on the energy market is mainly motivated by the rapid changes it has experienced.
Over the last decades, electricity markets have gone through an unprecedented transformation; this shift was driven by the liberalization of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis help also motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Interestingly, the process of deregulation lead to an increasing interest in the field of electricity price forecasting (EF) within the academic community \ref{fig:epf_evolution}.
\\
Moreover, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 SDGs \cite{un_sdgs}
\\
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
The are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low MAPE \ref{mape} results in increased savings \cite{savings}. Furthermore,the adoption of smart meters provides power market companies with a ton of consumer data; this will enable them to better model consumer preferences.
\\
Transmission system operators main goal is to match supply and deman, generally TSO do so by increasing or decreasing the generation supplied. Thus, from their view point forecasting is critical for balancing the electricity netwrok.
\\
Other possible applications are: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and hangling missing data.
\\
\section{Point vs probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either a prediction interval, quantiles or a probability distribution for each point in time. For this reason, probabilistic forecasts are more informative than point forecasts; and this is why the interest of the research community is shifting towards them.
Note that a probabilistic forecast can be turned into a point forecast by simply taking its expecatation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.

@@

- what is the question? objectives of the thesis
the description of the problem tackled and the methodology used to solve
it.

The scope of the thesis is analyzing state of the art forecasting methods in the energy market and to compare them with ideas coming from kernel theory. Applications of such methods to

>>>>

Individuals and organizations constantly face situations of uncertainty thus the need of robust forecasting methods. Such methods are crucial to the process of taking informed decision and to strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guesses about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting techniques to the energy sector. 
\\
The reason of our decision to focus on the energy market is mainly motivated by the rapid changes it has undergone. %/experienced 
Over the last decades, electricity markets have gone through an unprecedented transformation; this shift was driven by the liberalization of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis help also motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Furthermore, the process of deregulation lead to an increasing interest in the field of electricity price forecasting (EF) within the academic community \ref{fig:epf_evolution}.
\\
In addtion, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 SDGs \cite{un_sdgs}.
\\
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
The are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low MAPE \ref{mape} results in increased savings \cite{savings}. Furthermore,the adoption of smart meters provides power market companies with a ton of consumer data; this can enable them to better model consumer preferences.
\\
Transmission system operators' main goal is to match supply and deman, generally TSO do so by increasing or decreasing the generation supplied. Thus, from their view point forecasting is critical for balancing the electricity netwrok.
\\
Probabilistic forecasting may be useful to power producers, traders and consumers in order to improve their decision making process and managing risk(VaR). This holds in particular for traders, because probabilistic forecast enables them to simulate scenarios and carry out stress tests.
\\
Other possible applications are: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and handling missing data.
\\
\section{Point vs probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either a prediction interval, quantiles or a probability distribution for each point in time. For this reason, probabilistic forecasts are more informative than point forecasts; and this is why the interest of the research community is shifting towards them.
Note that a probabilistic forecast can be turned into a point forecast by simply taking its expecatation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.

@@

% - what is the question? objectives of the thesis
% the description of the problem tackled and the methodology used to solve
% it.

The scope of the thesis is analyzing state of the art forecasting methods in the energy market and to compare and to integrate them with ideas coming from the theory of kernel methods.
